---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Laila Abed"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

#Loading libraries and reading CVS processed datasets. 
library(tidyverse);library(agricolae);library(here)
here()
getwd()
raw_data <- 
  read.csv(here(
"./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"
), stringsAsFactors = TRUE)

# Convert the "sampledate" to date
raw_data$sampledate <- as.Date(raw_data$sampledate, format = "%m/%d/%y")
# Check if sample date is date now
class(raw_data$sampledate)

#2
# Define my custom ggplot theme
library(ggplot2)
custom_theme <- theme_minimal() +
theme(
text = element_text(size = 12, color = "black"),
axis.title = element_text(face = "bold"),
axis.text = element_text(color = "blue"),
panel.background = element_rect(fill = "lightgray"),
panel.grid.major = element_line(color = "gray"),
panel.grid.minor = element_blank()
)
# Set my custom theme as the default theme
theme_set(custom_theme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0:There is no significant change in mean lake temperature recorded during July 
with depth across all lakes. 
Ha: There is a significant change in mean lake temperature recorded during July 
with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
#Load the dplyr and tidyr packages
library(dplyr)
library(tidyr)
# Filter records with dates in July, select columns "lakename, year4, daynum, 
# depth, temperature_C", and remove NAs.
filtered_data <- raw_data %>%
filter(format(sampledate, "%m") == "07") %>%
select(lakename, year4, daynum, depth, temperature_C) %>%
drop_na()
# View the head of the filtered dataset
head(filtered_data)
#5
# Load the ggplot2 package
library(ggplot2)
# Create a scatter plot with a smoothed line
scatter_plot <- 
  ggplot(data = filtered_data, aes(x = depth, y = temperature_C)) +
geom_point(color = "blue", size = 3, alpha = 0.5) + 
geom_smooth(method = "lm", se = FALSE, color = "red") + 
# Customize the plot's appearance
labs(
title = "Scatter Plot for Temperature and Depth Relationship",
x = "Depth (m)",
y = "Temperature (C)"
) +
#theme_minimal() +
theme(
panel.grid.major = element_blank(),
panel.grid.minor = element_blank(),
axis.title = element_text(face = "bold"),
axis.text = element_text(size = 12),
plot.title = element_text(size = 16, hjust = 0.5)
) +
ylim(0, 35)
# Display the scatter plot
scatter_plot

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There is a negative linear relationship between depth and temperature.
As depth increases, the temperature tends to decrease.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# Linear regression model
linear_model <- lm(temperature_C ~ depth, data = filtered_data)
# Display the summary of the linear regression model
summary(linear_model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The linear regression model shows that there is a significant change 
in mean lake temperature recorded during July with depth across all lakes. 
Meaning changes in depth has a significant change in the lake temperature 
(negative relationship as depth increases, temperature decreases), accounting 
for nearly (R sequare) 74% of the variability (74% of the variability in 
temperature can be explained by changes in depth).  
The model is based on 9,726 data points (big sample size so it is 
very represenetaive to the population) and indicates a relationship between 
temperature and depth that is statistically significant. The p-values for both 
the intercept and slope of the depth are less than 0.05, indicating that these 
results are statistically significant at any conventional level. This suggests 
that for every 1-meter increase in depth, the temperature will decrease by 
approximately 1.946 degrees Celsius, as predicted by the estimated slope of the 
depth variable. Residual standard error: 3.835 on 9726 degrees of freedom. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
# Create a list of models with different combinations of predictor variables
models <- list(
model1 = lm(temperature_C ~ year4, data = filtered_data),
model2 = lm(temperature_C ~ daynum, data = filtered_data),
model3 = lm(temperature_C ~ depth, data = filtered_data),
model4 = lm(temperature_C ~ year4 + daynum, data = filtered_data),
model5 = lm(temperature_C ~ year4 + depth, data = filtered_data),
model6 = lm(temperature_C ~ daynum + depth, data = filtered_data),
model7 = lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)
)
model7 = lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)
step(model7)
# Calculate AIC for each model
AIC_values <- sapply(models, AIC)
# Find the model with the lowest AIC
best_model <- names(AIC_values)[which.min(AIC_values)]
# Display AIC values and the best model
AIC_values
#10
# Multiple regression model with year4, daynum, and depth as predictors
multiple_regression_model <- lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)
# Display the summary of the multiple regression model
summary(multiple_regression_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method uses variables (year4, daynum, and depth) to predict
temperature variability in the North Temperate Lakes LTER data. These variables 
provide the best predictors for explaining temperature changes, and the three of
them explain 74.12% of the varability of temperature. By comparing the model 
that uses all three variables with the model that uses only depth, we found that 
the model with all three variables provides an improvement in explaining 
temperature variability. This suggests that year4 and daynum provide additional 
information to the prediction of temperature beyond what can be explained by 
depth alone. The model including all three variables is better at explaining 
temperature varability than the model with only depth, which explains 73.87% of 
the temperature varability. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# ANOVA test to compare mean temperatures among lakes
anova_model <- aov(temperature_C ~ lakename, data = filtered_data)
# Display the summary of ANOVA
summary(anova_model)
# Fit a linear model with lakename as a predictor
linear_model_lakes <- lm(temperature_C ~ lakename, data = filtered_data)
# Display the summary of the linear model
summary(linear_model_lakes)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the 
lakes for July. Both the ANOVA test and the linear model provided evidence of 
the significant difference. The ANOVA results showed that the 
p-value is less than 0.05, which is statistically significant. This indicates 
that at least one lake has a different mean temperature from the others. 
Also, the linear model showed that the p-values for each lake 
(compared to the reference lake) are less than 0.05 indicating that the 
differences in mean temperatures for each lake are statistically significant.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

# Create the scatter plot with separate colors for each lake
ggplot(filtered_data, aes(x = depth, y = temperature_C, color = lakename)) +
geom_point(alpha = 0.5) +
geom_smooth(method = "lm", se = FALSE) +
ylim(0, 35) +
labs(
x = "Depth",
y = "Temperature (°C)",
title = "Temperature and Depth for Lakes in July",
color = "Lake Name"
) +
theme_minimal()

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

library(stats)
# Tukey's HSD test to compare means of different lakes
tukey_result <- HSD.test(aov(temperature_C ~ lakename, data = filtered_data),"lakename", group = T)
# Print
print(tukey_result)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:The results of the statistical analysis show that the temperature of 
Peter Lake is significantly different from Crampton Lake, East Long Lake, 
Hummingbird Lake, Tuesday Lake and West Long Lake. 
The adjusted p-value for the comparison between Peter Lake and Crampton
Lake is less than 0.05, indicating a statistically significant difference in 
their means. Also, the adjusted p-value for the comparison between Peter Lake 
and Hummingbird Lake is also less than 0.05, showing a statistically significant
difference in means. However, there is no statistically significant difference 
in means between Peter Lake and Paul Lake or between Peter Lake and Ward Lake. 
The adjusted p-value for the comparison between Peter Lake and Paul Lake; and 
Peter Lake and Ward Lake are greater than 0.05. This indicates that their means 
are not statistically different. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can do two samples t-test to compare the mean temperatures 
of Peter Lake and Paul Lake and determine whether they have distinct mean 
temperatures. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Subset the data for Crampton Lake and Ward Lake
crampton_lake_data <- filtered_data[filtered_data$lakename == "Crampton Lake", "temperature_C"]
ward_lake_data <- filtered_data[filtered_data$lakename == "Ward Lake", "temperature_C"]
# A two-sample t-test
t_test_result <- t.test(crampton_lake_data, ward_lake_data)
# Print the t-test result
print(t_test_result)
```

>Answer: Two-Sample t-test indicates: 1. The test does not provide
enough evidence to conclude that the mean temperatures for Crampton Lake and 
Ward Lake in July are significantly different. P-value is greater than 0.05. 
2. The null hypothesis that the mean temperatures of Crampton Lake and Ward Lake
are equal cannot be rejected based on the t-test. This result does not match the 
answer for part 16, which suggested that there were statistically significant 
differences in mean temperatures between lakes based on Tukey’s HSD test. 
In part 16, it was concluded that Peter Lake had the same mean temperature, 
statistically, as Paul Lake and Ward Lake. However, the t-test result 
for Crampton Lake and Ward Lake in part 18 suggests that their mean temperatures 
are not statistically distinct. The difference in results between part 16
and part 18 may be due to the different statistical tests used and their 
assumptions.
Tukey’s HSD test considers multiple pairwise comparisons among lakes, while the 
t-test focuses on a specific comparison between two lakes. 
Additionally, the t-test used a Welch correction to account for unequal 
variances, which can affect the results. These differences highlight the
importance of choosing appropriate statistical tests and considering the 
specific research question and data characteristics.
